FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-09-08 manuscript.Rmd
update 2024-09-08 08:51:02.338884

@@

The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

@@

When comparing common practice in contemporary psychological research to this idealized model, at first glance,
it appears that deductive research is overrepresented in the literature. 
An overwhelming 89.6% of published studies ostensibly tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged, or are forgotten [REF Meehl].

>>>>

The FAIR Guiding Principles (hereafter: FAIR principles) were initially established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since they were first introduced, scholars have demonstrated the relevance of these criteria for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
We introduce "FAIR theory", a digital representation of theory that can complies with the FAIR principles.

@@

The "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].

Concern that undisclosed flexibility in analyses was to blame for the abundance of false-positive findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
But have these reforms met their goal?
Recent reviews show that most preregistered hypothesis tests are not supported [@scheelExcessPositiveResults2021],
which suggests that the replication crisis may have been symptomatic of an underlying, and more fundamental, "theory crisis".
If psychological theories are insufficiently precise to derive testable hypotheses - then preregistration and replication will only serve to highlight those shortcomings.

@@

Contemporary psychological research falls short of this idealized model in several ways.
Firstly, because deductive research is overrepresented in the literature. 
According to one estimate, 89.6% of published studies tests hypotheses, which suggests that the literature is predominantly comprised of deductive research  [@kuhbergerPublicationBiasPsychology2014].
Closer examination reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies even reference specific theories [@mcphetresDecadeTheoryReflected2021].
Together, these statistics suggest the paradoxical role of theory in contemporary psychology.
Papers set out to test hypotheses, but these are rarely derived from theory, and test results rarely contribute to the improvement of existing theories.
Existing theories either persist unchanged for decades, or are forgotten [REF Meehl].
